Trajectories

### 2022-09-23 index.qmd
Fix random slopes and update manuscript

```{r}
#| label: prepare
#| echo: true
#| message: false
#| warning: false
# set params and load target contents

library(bayesplot)
library(brms) 
library(bayestestR)
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringdist)
library(stringr)
library(targets)
library(tibble)
library(tidybayes) 
library(tidyr)

@@

# Questionnaires

@@

# Items

@@

## Lexical frequency

@@

## \# Phonemes

@@

## Levenshtein

@@

## Frequency by \# Phonemes

@@

## Frequency by Levenshtein

@@

## \# Phonemes by Levenshtein

@@

# Participants

@@

## Age

@@

## Language profile

@@

## SES/parental education

```{r}
#| label: tbl-participants-edu
#| tbl-cap: Participant exposure to the language (*DoE*) and socio-economic status (*SES*)
#| echo: true
#| message: false
#| warning: false
participants %>%
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(multilex_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>% 
    summary_rows(
        columns = 2:7, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_number,
        decimals = 0
    ) %>% 
    sub_missing(columns = 2:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", 3:8) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:10)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

@@

# Data analysis

## Model comparison

@@

## Model equation

@@

## Model prior

@@

### Prior predictions

@@

## Model settings

@@

## R code (`brms`)

@@

### Model 0

@@

### Model 1

@@

### Model 2

@@

### Model 3

```{r}
print(model_fit_3$formula)
```
:::

## Stan code

Stan code generated by `brms::stancode`.

### Model 0
// generated with brms 2.16.1
functions {
/* compute correlated group-level effects
* Args: 
*   z: matrix of unscaled group-level effects
*   SD: vector of standard deviation parameters
*   L: cholesky factor correlation matrix
* Returns: 
*   matrix of scaled group-level effects
*/ 
matrix scale_r_cor(matrix z, vector SD, matrix L) {
// r is stored in another dimension order than z
return transpose(diag_pre_multiply(SD, L) * z);
}
/* cratio-logit log-PDF for a single response
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: ordinal thresholds
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
int nthres = num_elements(thres);
vector[nthres + 1] p;
vector[nthres] q;
int k = 1;
while (k <= min(y, nthres)) {
q[k] = log_inv_logit(disc * (mu - thres[k]));
p[k] = log1m_exp(q[k]);
for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
k += 1;
}
if (y == nthres + 1) {
p[nthres + 1] = sum(q);
}
return p[y];
}
/* cratio-logit log-PDF for a single response and merged thresholds
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: vector of merged ordinal thresholds
*   j: start and end index for the applid threshold within 'thres'
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
}
}
data {
int<lower=1> N;  // total number of observations
int Y[N];  // response variable
int<lower=2> nthres;  // number of thresholds
int<lower=1> K;  // number of population-level effects
matrix[N, K] X;  // population-level design matrix
// data for group-level effects of ID 1
int<lower=1> N_1;  // number of grouping levels
int<lower=1> M_1;  // number of coefficients per level
int<lower=1> J_1[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_1_1;
vector[N] Z_1_2;
vector[N] Z_1_3;
vector[N] Z_1_4;
vector[N] Z_1_5;
int<lower=1> NC_1;  // number of group-level correlations
// data for group-level effects of ID 2
int<lower=1> N_2;  // number of grouping levels
int<lower=1> M_2;  // number of coefficients per level
int<lower=1> J_2[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_2_1;
vector[N] Z_2_2;
vector[N] Z_2_3;
vector[N] Z_2_4;
vector[N] Z_2_5;
int<lower=1> NC_2;  // number of group-level correlations
int prior_only;  // should the likelihood be ignored?
}
transformed data {
int Kc = K;
matrix[N, Kc] Xc;  // centered version of X
vector[Kc] means_X;  // column means of X before centering
for (i in 1:K) {
means_X[i] = mean(X[, i]);
Xc[, i] = X[, i] - means_X[i];
}
}
parameters {
vector[Kc] b;  // population-level effects
vector[nthres] Intercept;  // temporary thresholds for centered predictors
vector<lower=0>[M_1] sd_1;  // group-level standard deviations
matrix[M_1, N_1] z_1;  // standardized group-level effects
cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
vector<lower=0>[M_2] sd_2;  // group-level standard deviations
matrix[M_2, N_2] z_2;  // standardized group-level effects
cholesky_factor_corr[M_2] L_2;  // cholesky factor of correlation matrix
}
transformed parameters {
real<lower=0> disc = 1;  // discrimination parameters
matrix[N_1, M_1] r_1;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_1] r_1_1;
vector[N_1] r_1_2;
vector[N_1] r_1_3;
vector[N_1] r_1_4;
vector[N_1] r_1_5;
matrix[N_2, M_2] r_2;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_2] r_2_1;
vector[N_2] r_2_2;
vector[N_2] r_2_3;
vector[N_2] r_2_4;
vector[N_2] r_2_5;
// compute actual group-level effects
r_1 = scale_r_cor(z_1, sd_1, L_1);
r_1_1 = r_1[, 1];
r_1_2 = r_1[, 2];
r_1_3 = r_1[, 3];
r_1_4 = r_1[, 4];
r_1_5 = r_1[, 5];
// compute actual group-level effects
r_2 = scale_r_cor(z_2, sd_2, L_2);
r_2_1 = r_2[, 1];
r_2_2 = r_2[, 2];
r_2_3 = r_2[, 3];
r_2_4 = r_2[, 4];
r_2_5 = r_2[, 5];
}
model {
// likelihood including constants
if (!prior_only) {
// initialize linear predictor term
vector[N] mu = Xc * b;
for (n in 1:N) {
// add more terms to the linear predictor
mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_1_4[J_1[n]] * Z_1_4[n] + r_1_5[J_1[n]] * Z_1_5[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n] + r_2_3[J_2[n]] * Z_2_3[n] + r_2_4[J_2[n]] * Z_2_4[n] + r_2_5[J_2[n]] * Z_2_5[n];
}
for (n in 1:N) {
target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);
}
}
// priors including constants
target += normal_lpdf(b[1] | 1, 0.1);
target += normal_lpdf(b[2] | 0, 0.1);
target += normal_lpdf(b[3] | 0, 0.1);
target += normal_lpdf(b[4] | 0, 0.1);
target += normal_lpdf(Intercept | -0.25, 0.1);
target += normal_lpdf(sd_1 | 1, 0.1)
- 5 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_1));
target += lkj_corr_cholesky_lpdf(L_1 | 2);
target += normal_lpdf(sd_2 | 1, 0.1)
- 5 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_2));
target += lkj_corr_cholesky_lpdf(L_2 | 2);
}
generated quantities {
// compute actual thresholds
vector[nthres] b_Intercept = Intercept + dot_product(means_X, b);
// compute group-level correlations
corr_matrix[M_1] Cor_1 = multiply_lower_tri_self_transpose(L_1);
vector<lower=-1,upper=1>[NC_1] cor_1;
// compute group-level correlations
corr_matrix[M_2] Cor_2 = multiply_lower_tri_self_transpose(L_2);
vector<lower=-1,upper=1>[NC_2] cor_2;
// additionally sample draws from priors
real prior_b_1 = normal_rng(1,0.1);
real prior_b_2 = normal_rng(0,0.1);
real prior_b_3 = normal_rng(0,0.1);
real prior_b_4 = normal_rng(0,0.1);
real prior_Intercept = normal_rng(-0.25,0.1);
real prior_sd_1 = normal_rng(1,0.1);
real prior_cor_1 = lkj_corr_rng(M_1,2)[1, 2];
real prior_sd_2 = normal_rng(1,0.1);
real prior_cor_2 = lkj_corr_rng(M_2,2)[1, 2];
// extract upper diagonal of correlation matrix
for (k in 1:M_1) {
for (j in 1:(k - 1)) {
cor_1[choose(k - 1, 2) + j] = Cor_1[j, k];
}
}
// extract upper diagonal of correlation matrix
for (k in 1:M_2) {
for (j in 1:(k - 1)) {
cor_2[choose(k - 1, 2) + j] = Cor_2[j, k];
}
}
// use rejection sampling for truncated priors
while (prior_sd_1 < 0) {
prior_sd_1 = normal_rng(1,0.1);
}
while (prior_sd_2 < 0) {
prior_sd_2 = normal_rng(1,0.1);
}
}

### Model 1
// generated with brms 2.16.1
functions {
/* compute correlated group-level effects
* Args: 
*   z: matrix of unscaled group-level effects
*   SD: vector of standard deviation parameters
*   L: cholesky factor correlation matrix
* Returns: 
*   matrix of scaled group-level effects
*/ 
matrix scale_r_cor(matrix z, vector SD, matrix L) {
// r is stored in another dimension order than z
return transpose(diag_pre_multiply(SD, L) * z);
}
/* cratio-logit log-PDF for a single response
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: ordinal thresholds
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
int nthres = num_elements(thres);
vector[nthres + 1] p;
vector[nthres] q;
int k = 1;
while (k <= min(y, nthres)) {
q[k] = log_inv_logit(disc * (mu - thres[k]));
p[k] = log1m_exp(q[k]);
for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
k += 1;
}
if (y == nthres + 1) {
p[nthres + 1] = sum(q);
}
return p[y];
}
/* cratio-logit log-PDF for a single response and merged thresholds
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: vector of merged ordinal thresholds
*   j: start and end index for the applid threshold within 'thres'
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
}
}
data {
int<lower=1> N;  // total number of observations
int Y[N];  // response variable
int<lower=2> nthres;  // number of thresholds
int<lower=1> K;  // number of population-level effects
matrix[N, K] X;  // population-level design matrix
// data for group-level effects of ID 1
int<lower=1> N_1;  // number of grouping levels
int<lower=1> M_1;  // number of coefficients per level
int<lower=1> J_1[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_1_1;
vector[N] Z_1_2;
vector[N] Z_1_3;
vector[N] Z_1_4;
vector[N] Z_1_5;
vector[N] Z_1_6;
int<lower=1> NC_1;  // number of group-level correlations
// data for group-level effects of ID 2
int<lower=1> N_2;  // number of grouping levels
int<lower=1> M_2;  // number of coefficients per level
int<lower=1> J_2[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_2_1;
vector[N] Z_2_2;
vector[N] Z_2_3;
vector[N] Z_2_4;
vector[N] Z_2_5;
int<lower=1> NC_2;  // number of group-level correlations
int prior_only;  // should the likelihood be ignored?
}
transformed data {
int Kc = K;
matrix[N, Kc] Xc;  // centered version of X
vector[Kc] means_X;  // column means of X before centering
for (i in 1:K) {
means_X[i] = mean(X[, i]);
Xc[, i] = X[, i] - means_X[i];
}
}
parameters {
vector[Kc] b;  // population-level effects
vector[nthres] Intercept;  // temporary thresholds for centered predictors
vector<lower=0>[M_1] sd_1;  // group-level standard deviations
matrix[M_1, N_1] z_1;  // standardized group-level effects
cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
vector<lower=0>[M_2] sd_2;  // group-level standard deviations
matrix[M_2, N_2] z_2;  // standardized group-level effects
cholesky_factor_corr[M_2] L_2;  // cholesky factor of correlation matrix
}
transformed parameters {
real<lower=0> disc = 1;  // discrimination parameters
matrix[N_1, M_1] r_1;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_1] r_1_1;
vector[N_1] r_1_2;
vector[N_1] r_1_3;
vector[N_1] r_1_4;
vector[N_1] r_1_5;
vector[N_1] r_1_6;
matrix[N_2, M_2] r_2;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_2] r_2_1;
vector[N_2] r_2_2;
vector[N_2] r_2_3;
vector[N_2] r_2_4;
vector[N_2] r_2_5;
// compute actual group-level effects
r_1 = scale_r_cor(z_1, sd_1, L_1);
r_1_1 = r_1[, 1];
r_1_2 = r_1[, 2];
r_1_3 = r_1[, 3];
r_1_4 = r_1[, 4];
r_1_5 = r_1[, 5];
r_1_6 = r_1[, 6];
// compute actual group-level effects
r_2 = scale_r_cor(z_2, sd_2, L_2);
r_2_1 = r_2[, 1];
r_2_2 = r_2[, 2];
r_2_3 = r_2[, 3];
r_2_4 = r_2[, 4];
r_2_5 = r_2[, 5];
}
model {
// likelihood including constants
if (!prior_only) {
// initialize linear predictor term
vector[N] mu = Xc * b;
for (n in 1:N) {
// add more terms to the linear predictor
mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_1_4[J_1[n]] * Z_1_4[n] + r_1_5[J_1[n]] * Z_1_5[n] + r_1_6[J_1[n]] * Z_1_6[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n] + r_2_3[J_2[n]] * Z_2_3[n] + r_2_4[J_2[n]] * Z_2_4[n] + r_2_5[J_2[n]] * Z_2_5[n];
}
for (n in 1:N) {
target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);
}
}
// priors including constants
target += normal_lpdf(b[1] | 1, 0.1);
target += normal_lpdf(b[2] | 0, 0.1);
target += normal_lpdf(b[3] | 0, 0.1);
target += normal_lpdf(b[4] | 0, 0.1);
target += normal_lpdf(b[5] | 0, 0.1);
target += normal_lpdf(Intercept | -0.25, 0.1);
target += normal_lpdf(sd_1 | 1, 0.1)
- 6 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_1));
target += lkj_corr_cholesky_lpdf(L_1 | 2);
target += normal_lpdf(sd_2 | 1, 0.1)
- 5 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_2));
target += lkj_corr_cholesky_lpdf(L_2 | 2);
}
generated quantities {
// compute actual thresholds
vector[nthres] b_Intercept = Intercept + dot_product(means_X, b);
// compute group-level correlations
corr_matrix[M_1] Cor_1 = multiply_lower_tri_self_transpose(L_1);
vector<lower=-1,upper=1>[NC_1] cor_1;
// compute group-level correlations
corr_matrix[M_2] Cor_2 = multiply_lower_tri_self_transpose(L_2);
vector<lower=-1,upper=1>[NC_2] cor_2;
// additionally sample draws from priors
real prior_b_1 = normal_rng(1,0.1);
real prior_b_2 = normal_rng(0,0.1);
real prior_b_3 = normal_rng(0,0.1);
real prior_b_4 = normal_rng(0,0.1);
real prior_b_5 = normal_rng(0,0.1);
real prior_Intercept = normal_rng(-0.25,0.1);
real prior_sd_1 = normal_rng(1,0.1);
real prior_cor_1 = lkj_corr_rng(M_1,2)[1, 2];
real prior_sd_2 = normal_rng(1,0.1);
real prior_cor_2 = lkj_corr_rng(M_2,2)[1, 2];
// extract upper diagonal of correlation matrix
for (k in 1:M_1) {
for (j in 1:(k - 1)) {
cor_1[choose(k - 1, 2) + j] = Cor_1[j, k];
}
}
// extract upper diagonal of correlation matrix
for (k in 1:M_2) {
for (j in 1:(k - 1)) {
cor_2[choose(k - 1, 2) + j] = Cor_2[j, k];
}
}
// use rejection sampling for truncated priors
while (prior_sd_1 < 0) {
prior_sd_1 = normal_rng(1,0.1);
}
while (prior_sd_2 < 0) {
prior_sd_2 = normal_rng(1,0.1);
}
}

### Model 2
// generated with brms 2.16.1
functions {
/* compute correlated group-level effects
* Args: 
*   z: matrix of unscaled group-level effects
*   SD: vector of standard deviation parameters
*   L: cholesky factor correlation matrix
* Returns: 
*   matrix of scaled group-level effects
*/ 
matrix scale_r_cor(matrix z, vector SD, matrix L) {
// r is stored in another dimension order than z
return transpose(diag_pre_multiply(SD, L) * z);
}
/* cratio-logit log-PDF for a single response
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: ordinal thresholds
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
int nthres = num_elements(thres);
vector[nthres + 1] p;
vector[nthres] q;
int k = 1;
while (k <= min(y, nthres)) {
q[k] = log_inv_logit(disc * (mu - thres[k]));
p[k] = log1m_exp(q[k]);
for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
k += 1;
}
if (y == nthres + 1) {
p[nthres + 1] = sum(q);
}
return p[y];
}
/* cratio-logit log-PDF for a single response and merged thresholds
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: vector of merged ordinal thresholds
*   j: start and end index for the applid threshold within 'thres'
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
}
}
data {
int<lower=1> N;  // total number of observations
int Y[N];  // response variable
int<lower=2> nthres;  // number of thresholds
int<lower=1> K;  // number of population-level effects
matrix[N, K] X;  // population-level design matrix
// data for group-level effects of ID 1
int<lower=1> N_1;  // number of grouping levels
int<lower=1> M_1;  // number of coefficients per level
int<lower=1> J_1[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_1_1;
vector[N] Z_1_2;
vector[N] Z_1_3;
vector[N] Z_1_4;
vector[N] Z_1_5;
vector[N] Z_1_6;
vector[N] Z_1_7;
int<lower=1> NC_1;  // number of group-level correlations
// data for group-level effects of ID 2
int<lower=1> N_2;  // number of grouping levels
int<lower=1> M_2;  // number of coefficients per level
int<lower=1> J_2[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_2_1;
vector[N] Z_2_2;
vector[N] Z_2_3;
vector[N] Z_2_4;
vector[N] Z_2_5;
int<lower=1> NC_2;  // number of group-level correlations
int prior_only;  // should the likelihood be ignored?
}
transformed data {
int Kc = K;
matrix[N, Kc] Xc;  // centered version of X
vector[Kc] means_X;  // column means of X before centering
for (i in 1:K) {
means_X[i] = mean(X[, i]);
Xc[, i] = X[, i] - means_X[i];
}
}
parameters {
vector[Kc] b;  // population-level effects
vector[nthres] Intercept;  // temporary thresholds for centered predictors
vector<lower=0>[M_1] sd_1;  // group-level standard deviations
matrix[M_1, N_1] z_1;  // standardized group-level effects
cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
vector<lower=0>[M_2] sd_2;  // group-level standard deviations
matrix[M_2, N_2] z_2;  // standardized group-level effects
cholesky_factor_corr[M_2] L_2;  // cholesky factor of correlation matrix
}
transformed parameters {
real<lower=0> disc = 1;  // discrimination parameters
matrix[N_1, M_1] r_1;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_1] r_1_1;
vector[N_1] r_1_2;
vector[N_1] r_1_3;
vector[N_1] r_1_4;
vector[N_1] r_1_5;
vector[N_1] r_1_6;
vector[N_1] r_1_7;
matrix[N_2, M_2] r_2;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_2] r_2_1;
vector[N_2] r_2_2;
vector[N_2] r_2_3;
vector[N_2] r_2_4;
vector[N_2] r_2_5;
// compute actual group-level effects
r_1 = scale_r_cor(z_1, sd_1, L_1);
r_1_1 = r_1[, 1];
r_1_2 = r_1[, 2];
r_1_3 = r_1[, 3];
r_1_4 = r_1[, 4];
r_1_5 = r_1[, 5];
r_1_6 = r_1[, 6];
r_1_7 = r_1[, 7];
// compute actual group-level effects
r_2 = scale_r_cor(z_2, sd_2, L_2);
r_2_1 = r_2[, 1];
r_2_2 = r_2[, 2];
r_2_3 = r_2[, 3];
r_2_4 = r_2[, 4];
r_2_5 = r_2[, 5];
}
model {
// likelihood including constants
if (!prior_only) {
// initialize linear predictor term
vector[N] mu = Xc * b;
for (n in 1:N) {
// add more terms to the linear predictor
mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_1_4[J_1[n]] * Z_1_4[n] + r_1_5[J_1[n]] * Z_1_5[n] + r_1_6[J_1[n]] * Z_1_6[n] + r_1_7[J_1[n]] * Z_1_7[n] + r_2_1[J_2[n]] * Z_2_1[n] + r_2_2[J_2[n]] * Z_2_2[n] + r_2_3[J_2[n]] * Z_2_3[n] + r_2_4[J_2[n]] * Z_2_4[n] + r_2_5[J_2[n]] * Z_2_5[n];
}
for (n in 1:N) {
target += cratio_logit_lpmf(Y[n] | mu[n], disc, Intercept);
}
}
// priors including constants
target += normal_lpdf(b[1] | 1, 0.1);
target += normal_lpdf(b[2] | 0, 0.1);
target += normal_lpdf(b[3] | 0, 0.1);
target += normal_lpdf(b[4] | 0, 0.1);
target += normal_lpdf(b[5] | 0, 0.1);
target += normal_lpdf(b[6] | 0, 0.1);
target += normal_lpdf(Intercept | -0.25, 0.1);
target += normal_lpdf(sd_1 | 1, 0.1)
- 7 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_1));
target += lkj_corr_cholesky_lpdf(L_1 | 2);
target += normal_lpdf(sd_2 | 1, 0.1)
- 5 * normal_lccdf(0 | 1, 0.1);
target += std_normal_lpdf(to_vector(z_2));
target += lkj_corr_cholesky_lpdf(L_2 | 2);
}
generated quantities {
// compute actual thresholds
vector[nthres] b_Intercept = Intercept + dot_product(means_X, b);
// compute group-level correlations
corr_matrix[M_1] Cor_1 = multiply_lower_tri_self_transpose(L_1);
vector<lower=-1,upper=1>[NC_1] cor_1;
// compute group-level correlations
corr_matrix[M_2] Cor_2 = multiply_lower_tri_self_transpose(L_2);
vector<lower=-1,upper=1>[NC_2] cor_2;
// additionally sample draws from priors
real prior_b_1 = normal_rng(1,0.1);
real prior_b_2 = normal_rng(0,0.1);
real prior_b_3 = normal_rng(0,0.1);
real prior_b_4 = normal_rng(0,0.1);
real prior_b_5 = normal_rng(0,0.1);
real prior_b_6 = normal_rng(0,0.1);
real prior_Intercept = normal_rng(-0.25,0.1);
real prior_sd_1 = normal_rng(1,0.1);
real prior_cor_1 = lkj_corr_rng(M_1,2)[1, 2];
real prior_sd_2 = normal_rng(1,0.1);
real prior_cor_2 = lkj_corr_rng(M_2,2)[1, 2];
// extract upper diagonal of correlation matrix
for (k in 1:M_1) {
for (j in 1:(k - 1)) {
cor_1[choose(k - 1, 2) + j] = Cor_1[j, k];
}
}
// extract upper diagonal of correlation matrix
for (k in 1:M_2) {
for (j in 1:(k - 1)) {
cor_2[choose(k - 1, 2) + j] = Cor_2[j, k];
}
}
// use rejection sampling for truncated priors
while (prior_sd_1 < 0) {
prior_sd_1 = normal_rng(1,0.1);
}
while (prior_sd_2 < 0) {
prior_sd_2 = normal_rng(1,0.1);
}
}

### Model 3
// generated with brms 2.16.1
functions {
/* compute correlated group-level effects
* Args: 
*   z: matrix of unscaled group-level effects
*   SD: vector of standard deviation parameters
*   L: cholesky factor correlation matrix
* Returns: 
*   matrix of scaled group-level effects
*/ 
matrix scale_r_cor(matrix z, vector SD, matrix L) {
// r is stored in another dimension order than z
return transpose(diag_pre_multiply(SD, L) * z);
}
/* cratio-logit log-PDF for a single response
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: ordinal thresholds
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_lpmf(int y, real mu, real disc, vector thres) {
int nthres = num_elements(thres);
vector[nthres + 1] p;
vector[nthres] q;
int k = 1;
while (k <= min(y, nthres)) {
q[k] = log_inv_logit(disc * (mu - thres[k]));
p[k] = log1m_exp(q[k]);
for (kk in 1:(k - 1)) p[k] = p[k] + q[kk];
k += 1;
}
if (y == nthres + 1) {
p[nthres + 1] = sum(q);
}
return p[y];
}
/* cratio-logit log-PDF for a single response and merged thresholds
* Args:
*   y: response category
*   mu: latent mean parameter
*   disc: discrimination parameter
*   thres: vector of merged ordinal thresholds
*   j: start and end index for the applid threshold within 'thres'
* Returns:
*   a scalar to be added to the log posterior
*/
real cratio_logit_merged_lpmf(int y, real mu, real disc, vector thres, int[] j) {
return cratio_logit_lpmf(y | mu, disc, thres[j[1]:j[2]]);
}
}
data {
int<lower=1> N;  // total number of observations
int Y[N];  // response variable
int<lower=2> nthres;  // number of thresholds
int<lower=1> K;  // number of population-level effects
matrix[N, K] X;  // population-level design matrix
// data for group-level effects of ID 1
int<lower=1> N_1;  // number of grouping levels
int<lower=1> M_1;  // number of coefficients per level
int<lower=1> J_1[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_1_1;
vector[N] Z_1_2;
vector[N] Z_1_3;
vector[N] Z_1_4;
vector[N] Z_1_5;
vector[N] Z_1_6;
vector[N] Z_1_7;
int<lower=1> NC_1;  // number of group-level correlations
// data for group-level effects of ID 2
int<lower=1> N_2;  // number of grouping levels
int<lower=1> M_2;  // number of coefficients per level
int<lower=1> J_2[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_2_1;
vector[N] Z_2_2;
vector[N] Z_2_3;
vector[N] Z_2_4;
vector[N] Z_2_5;
int<lower=1> NC_2;  // number of group-level correlations
int prior_only;  // should the likelihood be ignored?
}
transformed data {
int Kc = K;
matrix[N, Kc] Xc;  // centered version of X
vector[Kc] means_X;  // column means of X before centering
for (i in 1:K) {
means_X[i] = mean(X[, i]);
Xc[, i] = X[, i] - means_X[i];
}
}
parameters {
vector[Kc] b;  // population-level effects
vector[nthres] Intercept;  // temporary thresholds for centered predictors
vector<lower=0>[M_1] sd_1;  // group-level standard deviations
matrix[M_1, N_1] z_1;  // standardized group-level effects
cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
vector<lower=0>[M_2] sd_2;  // group-level standard deviations
matrix[M_2, N_2] z_2;  // standardized group-level effects
cholesky_factor_corr[M_2] L_2;  // cholesky factor of correlation matrix
}
transformed parameters {
real<lower=0> disc = 1;  // discrimination parameters
matrix[N_1, M_1] r_1;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_1] r_1_1;
vector[N_1] r_1_2;
vector[N_1] r_1_3;
vector[N_1] r_1_4;
vector[N_1] r_1_5;
vector[N_1] r_1_6;
vector[N_1] r_1_7;
matrix[N_2, M_2] r_2;  // actual group-level effects

>>>>

```{r}
#| label: prepare
#| echo: true
#| message: false
#| warning: false
# set params and load target contents
library(bayesplot)
library(brms) 
library(bayestestR)
library(conflicted)
library(dplyr) 
library(forcats)
library(ggplot2) 
library(ggsci)
library(gt) 
library(here) 
library(janitor)
library(knitr)
library(lubridate)
library(multilex)
library(patchwork) 
library(papaja) 
library(purrr)
library(scales) 
library(stringdist)
library(stringr)
library(targets)
library(tibble)
library(tidybayes) 
library(tidyr)

@@

# Methods

Data and materials are available at OSF (https://osf.io/hy984/), and code is available at GitHub (https://github.com/gongcastro/trajectories).

## Questionnaires

@@

## Items

@@

### Lexical frequency

@@

### \# Phonemes

@@

### Levenshtein

@@

### Frequency by \# Phonemes

@@

### Frequency by Levenshtein

@@

### \# Phonemes by Levenshtein

@@

## Participants

@@

### Age

@@

### Language profile

@@

### SES/parental education

```{r}
#| label: tbl-participants-edu
#| tbl-cap: Participant exposure to the language (*DoE*) and socio-economic status (*SES*)
#| echo: true
#| message: false
#| warning: false
participants %>%
    mutate(
        edu_parent = fct_explicit_na(as.factor(edu_parent)),
        doe_catalan = cut(doe_catalan, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
    ) %>% 
    count(doe_catalan, edu_parent) %>% 
    right_join(
        expand_grid(
            doe_catalan = levels(.$doe_catalan),
            edu_parent = unique(multilex_data$logs$edu_parent)
        )
    ) %>% 
    rename(total = n) %>% 
    pivot_wider(names_from = edu_parent, values_from = total, values_fill = 0) %>%
    mutate(total = rowSums(cbind(.[,3:8]), na.rm = TRUE)) %>% 
    rename(Total = total) %>% 
    select(-"NA") %>% 
    gt() %>% 
    cols_label(
        doe_catalan = "DoE Catalan"
    ) %>% 
    summary_rows(
        columns = 1:7, 
        fns = list(Total = ~sum(., na.rm = TRUE)),
        formatter = fmt_number,
        decimals = 0
    ) %>% 
    sub_missing(columns = 2:8, missing_text = "--") %>% 
    tab_spanner("Educational attainment", 3:8) %>% 
    tab_style(
        style = list(cell_borders(weight = px(2), sides = "left", color = "grey50")),
        locations = cells_body(columns = 9, rows = 1:10)
    ) %>% 
    tab_style(
        cell_text(style = "italic"),
        cells_column_labels(columns = 2:7)
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_labels(columns = c(1, 8))
    ) %>% 
    tab_style(
        cell_text(weight = "bold"),
        cells_column_spanners(spanners = "Educational attainment")
    )
```

@@

## Data analysis

### Model comparison

@@

### Model equation

```{mermaid}
//| label: fig-causal-diagram
//| fig-cap: Causal diagram
flowchart LR
  B(Age) --> A("P(Acquisition)")
  C(Exposure) --> A
  D(Frequency) --> A
  E(Phonemes) --> A
  F(Levenshtein) --> C
  F(Levenshtein) --> A
  A("P(Acquisition)") --> G[Understands/Says response]
```

@@

### Model prior

@@

#### Prior predictions

@@

### Model settings

@@

### R code (`brms`)

@@

@@

@@

@@

```{r}
print(model_fit_3$formula)
```

#### Model 4

```{r}
print(model_fit_4$formula)
```
:::
